---
title: "Lecture 2: Preliminaries and One-sample problem"
shorttitle: "STATS 205 Lecture 2"
author: "Pratheepa Jeganathan"
date: "04/05/2019"
output: 
  beamer_presentation:
    colortheme: "seahorse"
    slide_level: 2
    includes:
      in_header: header.tex
bibliography: nonparamterics.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 3,fig.height = 3,message=FALSE, warning=FALSE)
```

# Examples

## Example 1.7 (Spatial Ability Scores of Students)
- Data on a student’s spatial ability using four tests of visualization.
- For each student, a single score representing their overall measure of spatial ability. 
- The spatial ability scores for 68 female and 82 male
high school students enrolled in advanced placement calculus classes in Florida.
    - What is the distribution of spatial ability scores
for the population represented by this sample of data?
    - Does the distribution for the male students appear to possess different characteristics than that of the female students? 
- These questions are problems in density estimation

## Example 1.8 (Sunspots)
-  Data on mean monthly sunspot observations
collected at the Swiss Federal Observatory in Zurich and the Tokyo Astronomical Observatory from the years 1749 to 1983.
- Excessive variability over time, obscuring any underlying trend in the cycle of sunspot appearances.
- No apparent analytical form or simple parametric model.
- Powerful method for obtaining the trend from a noise in this case is wavelet estimation and thresholding.

#  Preliminaries

## Notations

- X: random variable
- x: realizations (observed random variables)
- $f(x)$: probability density function (pdf)
- $F_{X}(x) =  P(X \leq x )$: cumulative distribution function (cdf)
- $X_{1}, \cdots\ X_{n}$: random sample (independent and identically distributed)

## Distribution-free test statistic

- Test statistic: $T\left(\cdot\right) = T\left(X_{1}, \cdots\ X_{n}\right)$, function of the data.
    - Example: $T = \dfrac{\bar{X} -\mu}{s/\sqrt{n}}$, where $\bar{X} = \dfrac{\sum_{i=1}^{n}X_{i}}{n}$ and $s^2 = \dfrac{\sum_{i=1}^{n} \left(X_{i}-\bar{X}\right)^2}{n-1}$, $\mu$ is known under $\text{H}_{0}$.
    
-  Distribution-free test statistic
    - Example: $\mathcal{U} = \text{MVN}\left( \vmu = \left(\mu, \cdots, \mu\right), \mSigma = \sigma^{2}\mI\right)$
        - $T_{1} = \dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim \text{N}\left(0,1\right)$.
        - $T_{2} = \dfrac{\bar{X} - \mu}{s/\sqrt{n}} \sim \text{t}_{n-1}$.
        
- Nonparametric distribution-free test statistic
    - The class $\mathcal{U}$, $T\left(\cdot\right)$ is distribution free over contains more than one distributional forms.
    
##
-   Distribution-free confidence interval, distribution-free multiple comparison procedure, distribution-free confidence band, asymptotically distribution-free test statistic, asymptotically distribution-free multiple comparison procedure, and asymptotically distribution-free confidence band.

## Rank statistic

- Absolute rank: For any random variable $Z_{1}, \cdots\ Z_{n}$, the absolute rank of $Z_{i}$, denoted by $R_{i}$ is the rank of $|Z_{i}|$ among $|Z_{1}|, \cdots, |Z_{n}|$.
- Rank statistic: A statistic $T\left(\vR\right)$ based only on the ranks of a sample is rank statistic. 
    - $T\left(\vR\right)$ is distribution-free over iid joint continuous distribution.
- Signed rank: The signed rank of $Z_{i}$ is $R_{i}\psi_{i}$, where 
\begin{equation}
\psi_{i} = \left \lbrace  
\begin{array}{ll} 
1, & \text{if} \hspace{.5in} Z_{i} > 0,\\
0, & \text{if} \hspace{.5in} Z_{i} < 0.
\end{array}
\right.
\end{equation}
- Signed rank statistic: A statistic $T\left(\vpsi, \vR \right) = T\left(R_{1}\psi_{1}, \cdots, R_{n}\psi_{n} \right)$ that is a function of $Z_{1}, \cdots\ Z_{n}$ only through the signed ranks is the signed rank statistic.
    - $T\left(\vpsi, \vR \right)$ is ditribution-free over iid joint continuous distribution symmetric about 0.
 

# Sign test (Fisher) - paired replicates data/one-sample data

## Sign test 

- $Z_{1}, \cdots\ Z_{n}$ random sample from a continuous population that has a common median $\theta$.
    - If $Z_{i} \sim F_{i}$, $F_{i}\left(\theta\right) = F_{i}\left(Z_{i} \leq \theta \right) = F_{i}\left(Z_{i} > \theta \right) = 1- F_{i}\left(\theta\right)$.
- Hypothesis testing: 
    - $\text{H}_{0}: \theta = 0$ versus $\text{H}_{A}: \theta \neq 0$.

## Sign test (Cont.)

- Sign test statistic: $B = \sum_{i=1}^{n}\psi_{i}$.
- Motivation:
    - When $\theta$ is larger than 0, there will be larger number of positive $Z_{i}$s $->$ big $B$ value $->$ reject $\text{H}_{0}$ in favor of $\theta > 0$.
- Under $\text{H}_{0}$, $B \sim \left(n, 1/2 \right)$
- Significance level $\alpha$: probability of rejecting $H_{0}$ when it is true.
- Note
    - choices of $\alpha$ are limited to possible values of the $B \sim \left(n, 1/2 \right)$ cdf.
    - compare the distribution of $B$ under $\text{H}_{0}$ and the observed test statistic value.


## Sign test (Cont.)

- Rejection regions
    - $\text{H}_{A}: \theta > 0$, Reject $\text{H}_{0}$ if $B \geq b_{\alpha; n, 1/2}$.
    - $\text{H}_{A}: \theta < 0$, Reject $\text{H}_{0}$ if $B \leq n-b_{\alpha; n, 1/2}$.
    - $\text{H}_{A}: \theta \neq 0$, Reject $\text{H}_{0}$ if $B \geq b_{\alpha/2; n, 1/2}$ or $B \leq n-b_{\alpha/2; n, 1/2}$.

## Large-Sample Approximation (Sign test)

- $B^{*} = \dfrac{ B- \E_{0}\left(B\right) }{ \V_{0}\left(B\right)^{1/2} } \sim \text{N}\left(0,1\right) \hspace{.2in} \text{as} \hspace{.2in} n 	\to \infty \hspace{.2in} , \text{where}$
- $\E_{0}\left(B\right) = \dfrac{n}{2}$ and $\V_{0}\left(B\right)= \dfrac{n}{4}$

- Rejection regions
    - $\text{H}_{A}: \theta > 0$, Reject $\text{H}_{0}$ if $B^{*} \geq z_{\alpha}$.
    - $\text{H}_{A}: \theta < 0$, Reject $\text{H}_{0}$ if $B^{*} \leq -z_{\alpha}$.
    - $\text{H}_{A}: \theta \neq 0$, Reject $\text{H}_{0}$ if $B^{*} \geq z_{\alpha/2}$ or $B \leq -z_{\alpha/2}$.
    

## Ties  (Sign test)

- Discard zero $Z$ values and redefine $n$.
- If too many zeros, choose alternative statistical procedure (Chapter 10)




<!-- - Order statistics facts: $X_{1}, \cdots\ X_{n} \sim F(x)$, $F(X_{(1)}) \leq  \cdots \leq F(X_{(n)})$ -->
<!-- - probability integral transfomation -->

<!-- - Distribution of ranks -->
<!-- - Symmetric distribution: $\left( X-\mu\right) \overset{d}{=} \left( \mu - X\right)$ -->

<!-- - Homework problem 1: Assume $X$ is symmetric about $\mu$. Show that $F(\mu+t) =  1- F(\mu-t)$ (definition of symmetry). -->

<!-- - Statistic $t(\cdot)$ -->
<!--     - translational statistic -->
<!--     - translational invariant statistic -->

<!-- - Homework problem 2: Show that $\bar{X}$ mean is translational statistic and $S^{2}$ sample variance is translational invariant statistic. -->

<!-- Example (HWC: Chapter 3, Example 3.2, pg. 44) -->

<!-- - Government and private workers were matched. -->
<!-- - Salaries of matched pairs were obtained. -->
<!-- - $X =$ government worker's salary and  $Y =$ matched private sector salary. -->
<!-- - Test that government workers are generally paid less than their counterparts in the private sector. -->

```{r rej-region, echo=FALSE, message=FALSE, warnings=FALSE, results='asis', include=FALSE}
#tabl = "
# |$\\text{H}_{A}$  |Reject $\\text{H}_{0}$ is          |
# |-----------------|-----------------------------------|
# |$\\text{H}_{A}: \\theta > 0$ |$B \\geq b_{\\alpha; n, 1/2}$|
# |$\\text{H}_{A}: \\theta < 0$|$B \\leq n-b_{\\alpha; n, 1/2}$|
# |$\\text{H}_{A}: \\theta \\neq 0$| $B \\geq b_{\\alpha/2; n, 1/2}$ or $B \\leq n-b_{\\alpha/2; n, 1/2}$|
# "
# cat(tabl)
```

```{r echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
# sign.rank.test = "
# |$\\text{H}_{A}$  |Reject $\\text{H}_{0}$ is          |
# |-----------------|-----------------------------------|
# |$\\text{H}_{A}: \\theta > 0$ |$T^{+} \\geq t_{\\alpha}$|
# |$\\text{H}_{A}: \\theta < 0$|$T^{+} \\leq \dfrac{n(n+1)}{2}-t_{\\alpha}$|
# |$\\text{H}_{A}: \\theta \\neq 0$| $T^{+} \\geq t_{\\alpha/2}$ or $T^{+} \\leq \dfrac{n(n+1)}{2}-t_{\\alpha/2}$|
# "
# cat(sign.rank.test)
```

<!-- ## Boxplot to check the symmetry -->

```{r eval=FALSE, include=FALSE}
library(tidyr)
library(ggplot2)
library(magrittr)
df.long = gather(df, key = "type")
df.long = df.long %>% data.frame
df.long$type = factor(df.long$type)
ggplot(df.long, aes(x = type, y = value)) +
  geom_boxplot()
```

<!-- ##  -->

```{r echo=FALSE, include=FALSE, eval=FALSE}
library(tidyr)
library(ggplot2)
library(magrittr)
df.long = gather(df, key = "type")
df.long = df.long %>% data.frame
df.long$type = factor(df.long$type)
ggplot(df.long, aes(x = type, y = value)) +
  geom_boxplot()
```

<!-- ## -->
```{r eval=FALSE, include=FALSE}
ggplot(df, aes(y = Z)) +
  geom_boxplot()
```

<!-- ## -->
```{r echo=FALSE, include=FALSE,eval=FALSE}
ggplot(df, aes(y = Z)) +
  geom_boxplot()
```

```{r echo=FALSE, message=FALSE, warnings=FALSE, results='asis', include=FALSE}
# lar.sam.app.sign.test <- "
# |$\\text{H}_{A}$  |Reject $\\text{H}_{0}$ is          |
# |-----------------|-----------------------------------|
# |$\\text{H}_{A}: \\theta > 0$ |$B^{*} \\geq z_{\\alpha}$|
# |$\\text{H}_{A}: \\theta < 0$|$B^{*} \\leq z_{\\alpha}$|
# |$\\text{H}_{A}: \\theta \\neq 0$| $B^{*} \\geq z_{\\alpha/2}$ or $B \\leq z_{\\alpha/2}$|
# "
# cat(lar.sam.app.sign.test)
```


# References

##  References for this lecture
HWC: Chapter 1.2

HWC: Chapter 1.3

HWC: Chapter 3.4-3.6


