---
title: "Lecture 9: Two-sample problem I"
shorttitle: "STATS 205 Lecture 9"
author: "Pratheepa Jeganathan"
date: "04/22/2019"
output: 
  beamer_presentation:
    colortheme: "seahorse"
    slide_level: 2
    includes:
      in_header: header.tex
bibliography: nonparamterics.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 3,fig.height = 3, fig.align="center",message=FALSE, warning=FALSE)
```

#  Recall

##
- One sample sign test, Wilcoxon signed rank test, large-sample approximation, median, Hodges-Lehman estimator, distribution-free confidence interval.
- Jackknife for bias and standard error of an estimator.
- Bootstrap samples, bootstrap replicates.
- Bootstrap standard error of an estimator.
- Bootstrap percentile confidence interval.
- Hypothesis testing with the bootstrap (one-sample problem.)
- Assessing the error in bootstrap estimates.
- Example: inference on ratio of heart attack rates in the aspirin-intake group to the placebo group.
- The exhaustive bootstrap distribution
- Discrete data problems (one-sample, two-sample proportion tests, test of homogeneity, test of independence)


## Two-sample location problem

- Data: two samples from two populations. $X_{1}, \cdots X_{m}$ from population 1 and corresponding to distribution function $F\left(.\right)$ and $Y_{1}, \cdots Y_{n}$ and corresponding to distribution function $G\left(.\right)$.
- Assumptions:
    - $X$ and $Y$ are mutually independent.
    - $F\left(.\right)$ and $G\left(.\right)$ are continuous.
    - Populations differ only by a location shift.
- Hypothesis 
    - $\text{H}_{0}: F\left(t\right) = G\left(t\right)$ for each $t$ versus $\text{H}_{A}: \Delta \neq 0$, where $G\left(t\right) = F\left(t - \Delta\right)$.
    - Other alternatives $\text{H}_{A}: \Delta > 0$, $\text{H}_{A}: \Delta < 0$ (stochastic ordering of alternatives).
    - For example, $\Delta$ can be expected treatment effect compared to control.

## Two-sample location problem

- Wilcoxon two-sample rank sum test
    - Combined sample of $N = m+n$.
    - Let $S_{n}$ denote the rank of $Y_{n}$ in this joint ordering.
    - Wilcoxon two-sample rank sum statistic $W = \sum_{j=1}^{n}S_{j}$.
- R command: `wilcox.test(x, y, paired = FALSE, alternative = "two.sided")`.
- Under $\text{H}_{0}$, all $\binom{N}{n}$ possible assignments for the $Y$-ranks are equally likely, each having probability $\dfrac{1}{\binom{N}{n}}$.

## Two-sample location problem
- Mann–Whitney statistic
    - $U = \sum_{i=1}^{m}\sum_{j=1}^{n}\phi\left(X_{i}, Y_{j} \right)$, where $\phi\left(X_{i}, Y_{j} \right) = 1 \hspace{.1in} if \hspace{.1in} X_{i} < Y_{j}$.
- Tests based on $W$ and $U$ are equivalent.
    - $W = U + \dfrac{n(n+1)}{2}$.
    
## Two-sample location problem (large-sample approximation)

- Test statistic: the standarized version of $W$.
    - $Z = \dfrac{W - \lbrace m \left(m+n+1\right)/2 \rbrace}{\lbrace mn\left(m+n+1\right)/12\rbrace ^{1/2}}$.
    - $Z \sim \text{N}\left(0, 1\right)$.
- If there are ties
    - Use the average of the ranks.
    - not an exact test.

## Example (Two-sample location problem)
- **HWC** Example 4.1
- Whether there is a difference in the transfer of tritiated water (water containing tritium, a radioactive isotope of hydrogen) across the tissue layers in the term human chorioamnion (a placental membrane) and in the human chorioamnion between 3- and 6-months’ gestational age.
- Data: measurd the permeability constant Pd of the human chorioamnion to water. 
    - collected data after 5 min of delivery from the placentas of healthy subjects, between 12 and 26 weeks following termination of pregnancy and 
    - from term, uncomplicated vaginal deliveries.
- $\text{H}_{0}: \Delta = 0$ versus $\text{H}_{A}: \Delta \neq 0$.

## Example (Two-sample location problem)
```{r}
at.term = c(.80, .83, 1.89, 1.04, 1.45, 
  1.38, 1.91, 1.64, .73, 1.46)
gest.age = c(1.15, .88, .90, .74, 1.21)
wilcox.test(at.term, gest.age, 
  alternative="greater", conf.int=T)
```

We do not have enough evidencce to reject the $\text{H}_{0}$.

## Two-sample location problem (Hodges–Lehmann estimator)

- Compute $mn$ differences $Y_{j}- X_{i}; i = 1, \cdots, m, j = 1, \cdots, n$. $\hat{\Delta} = \text{median}\lbrace Y_{j}- X_{i}; i = 1, \cdots, m, j = 1, \cdots, n\rbrace$.
- `wilcox.test` gives a point estimator: `difference in location`
- In our example $\hat{\Delta} = .305$.
- There is no need to perform the large-sample approximation because we have the result for the exact test.

## Comments on Wilcoxon rank sum test

- The confidence interval for $\Delta$ is found directly from the R command wilcox.test.
- The significance level of the rank sum test is not preserved if the two populations differ in dispersion or shape.
- If $\text{H}_{0}: \Delta = \Delta_{0}$, form a pseudo-sample $Y^{'}_{1} = Y_{1} - \Delta_{0}, \cdots, Y^{'}_{n} = Y_{n} - \Delta_{0}$.
- Run Wilcoxon rank sum test on $X_{1}, \cdots, X_{m}, Y^{'}_{1}, \cdots, Y^{'}_{n}$.
- In a two-sample problem, we sometimes wish to estimate $\delta = P\left( X < Y\right)$. Read **HWC** page 138 - 141, comment 18.
    - For example in quantifying stress-strength reliability.
    - $X$ stress (random) applied to a system.
    - $Y$ strength of the system.

##  References for this lecture

**HWC** Chapter 4.1-4.3

<!-- Homework problem: Page 133, problem 1 (rank sum test - large-sample approximation) -->

<!-- Homework problem: Page 134, problem 5 (rank sum test - Wilcoxon) -->

<!-- Homework problem: Page 141, problem 21 (estimate $\delta$ and confidence interval for $\delta$). -->