---
title: "Lecture 4: Statistical functionals and Influence functions"
shorttitle: "STATS 205 Lecture 4"
author: "Pratheepa Jeganathan"
date: "04/10/2019"
output: 
  beamer_presentation:
    colortheme: "seahorse"
    slide_level: 2
    includes:
      in_header: header.tex
bibliography: nonparamterics.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 3,fig.height = 3,message=FALSE, warning=FALSE)
```


# Robustness

## Properties of estimators

- Measures of robustness
    - efficiency
    - influence
    - breakdown 
- Asymptotic relative efficiency **HWC** Chapter 3.11

- Consider influence and breakdown


##  Sensitivity to gross errors

- Sensitivity curve: function of observations.
- Let $\vz_{n} = \left(z_{1}, \cdots, z_{n}\right)^{T}$ drawn from cdf $F$ and $\theta$ is the location parameter.
- Let $\hat{\theta}$ is an estimator of $\theta$.
- Add an outlier observation $z$ to $\vz_{n}$, $\vz_{n+1} = \left(z_{1}, \cdots, z_{n}, z\right)^{T}$.
- The sensitivity curve of an estimator $\hat{\theta}$ is
\begin{equation}
S\left(z; \hat{\theta}\right) = \dfrac{\hat{\theta}_{n+1} - \hat{\theta}_{n}}{1/(n+1)}.
\end{equation}

## Sensitivity to gross errors (examples)

```{r}
z_n = c(1.85, 2.35, -3.85, -5.25, -0.15, 
  2.15, 0.15, -0.25, -0.55, 2.65)
mean(z_n)
```
```{r}
median(z_n)
```

```{r}
library(ICSNP)
hl.loc(z_n)
```

## Example (Sensitivity curve for mean)
```{r}
z_n_plus_1_df = data.frame(z_n_plus_1 = seq(-20, 20, 
  by = 1))

sensitivity <- function(theta_n_plus_1, theta_n, n){
  (theta_n_plus_1- theta_n)*(n+1)
}

mean_z_n_plus_1 = apply(z_n_plus_1_df, 1, function(x){
  x = c(z_n,x)
  mean(x)
})

sensitivity_mean = sensitivity(mean_z_n_plus_1, 
  mean(z_n), length(z_n))
```

## Example (Sensitivity curve for median)

```{r}
median_z_n_plus_1 = apply(z_n_plus_1_df, 1, function(x){
  x = c(z_n,x)
  median(x)
})

sensitivity_median = sensitivity(median_z_n_plus_1, 
  mean(z_n), length(z_n))
```

## Example (Sensitivity curve for HL)

```{r}
HL_z_n_plus_1 = apply(z_n_plus_1_df, 1, function(x){
  x = c(z_n,x)
  hl.loc(x)
})
sensitivity_HL = sensitivity(HL_z_n_plus_1, 
  mean(z_n), length(z_n))
```

## Example (Sensitivity curve)

```{r fig.show = 'hide'}
library(tidyr)
library(ggplot2)
df = data.frame(z_n_plus_1 = z_n_plus_1_df$z_n_plus_1, 
  sensitivity_mean = sensitivity_mean, 
  sensitivity_median = sensitivity_median, 
  sensitivity_HL = sensitivity_HL)
df_long = gather(df, key = "estimator", 
  value = "value", -z_n_plus_1)
df_long$estimator = factor(df_long$estimator)
ggplot(data = df_long) + 
  geom_line(aes(x = z_n_plus_1, 
    y = value, group =estimator, color = estimator)) + 
  xlab(bquote(z[n+1]))+ 
  scale_color_discrete(name = "Estimator", 
    labels = c("HL", "Mean", "Median")) + ylab(bquote(S(z[n+1],hat(theta))))
```

## Example (Sensitivity curve)

- Mean: unbounded.
- Median and Hodges–Lehmann: bounded.
```{r echo=FALSE, fig.width=5.5}
ggplot(data = df_long) + 
  geom_line(aes(x = z_n_plus_1, 
    y = value, group =estimator, color = estimator)) + 
  xlab(bquote(z[n+1]))+ 
  scale_color_discrete(name = "Estimator", 
    labels = c("HL", "Mean", "Median")) + ylab(bquote(S(z[n+1],hat(theta))))
```


## Statistical functionals

- Statistical inference involves estimating some aspects of a cdf $F$ on the basis of a random sample drawn from $F$.
- Statistical functional $T\left(F\right)$: any function of $F$.
    - Let $Z_{1}, \cdots, Z_{n} \sim F$, where $F\left(z\right) = P\left(Z \leq z\right)$, define $\theta = T\left(F\right)$. 
- Examples:
    - Mean: $T\left(F\right) = \int z dF\left(z\right)$.
    - Median: $T\left(F\right) = F^{-1}\left(1/2\right)$.
    - HL: $T\left(F\right) = (1/2)\lbrace F * F \rbrace^{-1}(1/2),$ where $*$ denotes convolution.


##  Estimating statistical functionals

- Estimator of $F$: empirical CDF $\hat{F}\left(z\right) = \dfrac{\#\lbrace z_{i} \leq z\rbrace}{n}$.
- Plug-in principal: plug-in estimator of $T\left(F\right)$ is $T\left(\hat{F}\right)$ - (summary statistic).
- Plug-in principal is good when there is information about $F$ only through sample $\vz$ (not from the model). 


## Influence functions

- Influence function
    
    - Measures rate of change of $T\left(F\right)$ under small contamination at $z$ (kind of derivative).
    - Indicates statistical accuracy of a statistic (if influence function is bounded - robustness).
    - Useful for computing the approximate standard error of plug-in estimate $T\left(\hat{F}\right)$ (standard deviation of a summary statistic).
- Gateaux derivative of $T$ at $F$ in the direction $G$
    \begin{equation}
    L\left(G\right) = \lim_{\epsilon \rightarrow 0} \dfrac{T\left( (1-\epsilon)F + \epsilon G\right) - T\left(F\right)}{\epsilon}
    \end{equation}
    
## Influence functions
- If $G = \delta_{z}$ is a point mass at $z$
     \begin{equation}
    L\left(z\right) = \lim_{\epsilon \rightarrow 0} \dfrac{T\left( (1-\epsilon)F + \epsilon \delta_{z}\right) - T\left(F\right)}{\epsilon}.
    \end{equation}
- $L\left(z\right)$ is the influence function.
    
- Empirical influence function/plug-in estimator for $L\left( z\right)$
    \begin{equation}
    \hat{L}\left(z\right) = \lim_{\epsilon \rightarrow 0} \dfrac{T\left( (1-\epsilon)\hat{F} + \epsilon \delta_{z}\right) - T\left(\hat{F}\right)}{\epsilon}.
    \end{equation}
    
## Examples (influence functions)

- The influence function for our estimators are (up to constant of proportionality and center)
    - Mean: $z$
    - Median: $\text{sign}\left(z\right)$
    - HL: $F\left(z\right) - .5$
    
- Mean is not robust, but median and HL are robust

## Example (influence curves)
```{r}
influence_mean = z_n
influence_median = sign(z_n)
z_n_df = data.frame(z_n = z_n)
influence_HL = apply(z_n_df, 1, function(x){
  mean(z_n <= x) -.5
})
```

## Example (influence curves)
```{r fig.show='hide'}
df_inf = data.frame(z = z_n, 
  influence_mean = influence_mean, 
  influence_median = influence_median, 
  influence_HL = influence_HL)
df_inf_long = gather(df_inf, key = "estimator", 
  value = "value", -z)
df_inf_long$estimator = factor(df_inf_long$estimator)
ggplot(data = df_inf_long) + 
  geom_line(aes(x = z, 
    y = value, group =estimator, color = estimator)) + 
  xlab("z")+ 
  scale_color_discrete(name = "Estimator", 
    labels = c("HL", "Mean", "Median")) +
  ylab("L(z)")
```

## Example (influence curves)
```{r echo=FALSE, fig.width=5.5}
ggplot(data = df_inf_long) + 
  geom_line(aes(x = z, 
    y = value, group =estimator, color = estimator)) + 
  xlab("z")+ 
  scale_color_discrete(name = "Estimator", 
    labels = c("HL", "Mean", "Median")) +
  ylab("L(z)")
```

## Standard error of a plug-in estimator

- If $T\left(F\right) = \int a\left(z\right) dF\left(z\right)$, a linear functional 
    - $L\left( z\right) = a\left(z\right) - T\left(F\right)$. 
    - $\E \left(L\left( z\right) \right) = 0$.
    - $\tau^{2} = \int L\left( z\right)^2 dF\left(z\right) = \int \left(a\left(z\right) - T\left(F\right) \right)^{2} dF\left(z\right)$.
    - $\hat{\tau}^{2} = \dfrac{1}{n}\sum_{i=1}^{n} \left(a\left(Z_{i}\right) - T\left(\hat{F}\right) \right)^{2}$.
    - $\text{se}^{2}\left(T\left(\hat{F}\right) \right) = \dfrac{\hat{\tau}^{2}}{n}$.

## Example (Standard error of a plug-in estimator)

- $\theta = T\left(F\right)  = \int z dF\left(z\right)$.
- $\hat{\theta} = T\left(\hat{F}\right) = \int z d\hat{F}\left(z\right) = \dfrac{\sum_{i=1}^{n} Z_{i}}{n} = \bar{Z}$.
- $L\left(z\right) = z - \int z dF\left(z\right)$.
- $\hat{L}\left( z\right) = z - \bar{Z}$.
- $\text{se}^{2}\left(T\left(\hat{F}\right) \right) = \dfrac{n^{-1}\sum_{i=1}^{n} \left(Z_{i} -\bar{Z} \right)^{2}}{n}$.

## Breakdown point of an estimator

- Reference: Following notes from this [link](http://www.stat.umn.edu/geyer/5601/notes/break.pdf).

- Suppose we contaminate $n-m$ points in our sample
  $$z_{n}^{*} = \left(z_{1}, \cdots, z_{m},z_{m+1}^{*}, \cdots, z_{n}^{*} \right)$$.

- Consider $z_{m+1}^{*}, \cdots, z_{n}^{*}$ are very large $(\text{close to} \hspace{.1in} \infty)$.

- Breakdown point: the smallest value $n-m$ so that $\hat{\theta}\left( z_{n}^{*}\right)$ is bad.

- Finite sample breakdown point : a function of sample size $\dfrac{n-m}{n}$. 

- Asymptotic breakdown point: (single number) the limit of the finite sample breakdown point as $n \to \infty$.


## Examples (breakdown point)

- Sample mean
    - Finite sample BP: $\dfrac{1}{n}$.
    - Asymptotic BP: 0.
- Sample median
    - Finite sample BP: $\left[\dfrac{n-1}{2n}\right]$, $\left[u\right]$ is the largest integer less than or equal to $u$.
    - Asymptotic BP: .5.
- HL
    - Finite sample BP: [Read this notes](http://www.stat.umn.edu/geyer/5601/notes/break.pdf).
    - Asymptotic BP: $\approx 0.29$.

# References

##  References for this lecture


**W** Chapter 2 (Statistical functionals and influence functions)

**ET** Chapter 4, 5, 21.3 (Statistical functionals and influence functions)

**KM** Chapter 3.5 (R codes for sensitivity, breakdown, influence)

**HWC** Chapter 3.2 page 57, comment 16 (sensitivity to gross errors- HL)

**HWC** Chapter 3.5 page 77, comment 40 (sensitivity to gross errors- median)

